Active-contours driven registration method for structure-informed segmentation of diffusion MR images

### 2015-04-07 Publications/2015-NeuroImage/03-introduction.tex
edits received from Andres
edits from Chus

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:introduction}
The accurate delineation of \gls*{wm} in \gls*{dmri} and the fusion of prior
  anatomical information extracted from a \gls*{t1} image of the same subject
  are crucial in a range of applications based on tractography, such as
  the extraction of structural connectivity \citep{craddock_imaging_2013} or
  tract-based spatial statistics \citep{smith_tractbased_2006}.
However, segmenting \gls*{dmri} data precisely and coregistration of anatomical
  images are difficult for several reasons.
First, \gls{dmri} images suffer from partial voluming of several tissues in 
  interfacing voxels due to the low resolution of images (generally around
  \isores[mm]{2.2}).
Second, \gls*{dmri} schemes probe the diffusion process within the brain in 
  many angles called \glspl*{dwi}, completed by one or more baseline (\emph{b0}) 
  volumes without directional gradients.
The extremely low \gls*{snr} and high dimensionality of \glspl*{dwi} disable their
  use for direct segmentation.
The low contrast between \gls*{gm} and \gls*{wm} in the \emph{b0} volumes is not suitable for 
  segmentation either.
Lastly, \gls*{dmri} images are acquired using \gls*{epi} to speed up acquisition
  at the cost of presenting geometrical distortion and 
  signal losses \citep{jezzard_correction_1995} that hinder the segmentation by
  registering an anatomical image as source of same-subject prior information.
The artifact has been proven to cause an important impact on the anatomy extracted
  from \gls*{dmri} \citep{irfanoglu_effects_2012}, particularly in certain fiber bundles
  when used in whole-brain tractography.

@@

Therefore, given the described circumstances, a method performing simultaneously
  segmentation in \gls*{dmri} space and registration of the corresponding \gls*{t1} image
  to the \gls*{dmri} data could provide the optimal solution.
Previously, joint segmentation and registration has been successfully applied to
  longitudinal object tracking \citep{paragios_level_2003} and atlas-based
  segmentation \citep{gorthi_active_2011}.
The most extended approach is integrating active contours in image registration 
  methods.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  improved an existing method \citep{yezzi_variational_2003} based on linear registration
  to the nonlinear case implementing a free-form deformation field.
\cite{droske_mumfordshah_2009} reviewed the existing techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method in 2D using levelsets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  levelset function for the registration of several active contours, in the application
  of atlas-based segmentation.

The hypothesis under this research is that the registration and segmentation
  problems in \gls*{dmri} can be solved simultaneously, increasing the geometrical
  accuracy of the process.
We propose a tool called \emph{regseg} to exploit the prior information of shapes
  extracted on \gls*{t1} images to register the anatomical reference
  to \gls*{dmri} space, implicitly segmenting these data.
The approach uses active contours without edges \citep{chan_active_2001} that evolve driving a
  free-form deformation field of B-Spline basis functions.
Optimization is performed using a descent strategy of explicit shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}, therefore \emph{regseg}
  does not implement levelsets as most of the presented methods.
Since the nonlinear distortion is aligned with one of the imaging axis (see 
  \autoref{sec:human_connectome}), \emph{regseg} includes an anisotropic regularizer of
  the displacements field proposed by \cite{nagel_investigation_1986}.
Finally, we evaluate \emph{regseg} using an extension of our instrumentation framework
  \citep{esteban_simulationbased_2014} that simulates known and realistic distortions
  on \gls*{dmri} data.
The evaluation includes a comparison to the \gls*{t2b} correction by integrating in the framework
  an in-house implementation of the method.

>>>>

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:introduction}
The accurate delineation of \gls*{wm} in \gls*{dmri} and the fusion of prior
  anatomical information extracted from a \gls*{t1} image of the same subject
  are crucial in a range of applications based on tractography, such as
  the extraction of structural connectivity \citep{craddock_imaging_2013} or
  tract-based spatial statistics \citep{smith_tractbased_2006}.
However, segmenting \gls*{dmri} data precisely and coregistration of anatomical
  images are difficult due to several limitations.
First, \gls{dmri} images suffer from partial voluming of several tissues in 
  voxels located at their interfaces due to the low resolution of images
  (generally around \isores[mm]{2.2}).
Second, \gls*{dmri} schemes probe the diffusion process within the brain in 
  many angles called \glspl*{dwi}, completed by one or more baseline (\emph{b0}) 
  volumes without directional gradients.
The extremely low \gls*{snr} and high dimensionality of \glspl*{dwi} disable their
  direct use in segmentation.
The low contrast between \gls*{gm} and \gls*{wm} in the \emph{b0} volumes is not suitable for 
  segmentation either.
Lastly, \gls*{dmri} images are acquired using \gls*{epi} to speed up acquisition
  at the cost of presenting geometrical distortion and signal degradation and
  destruction \citep{jezzard_correction_1995}.
The artifact has been proven to cause an important impact on the anatomy extracted
  from \gls*{dmri} \citep{irfanoglu_effects_2012}, particularly in certain fiber bundles
  when used in whole-brain tractography.
These limitations preclude segmentation in native \gls*{dmri} space, including
  registration-based approaches using anatomical images as source of structural
  information.

@@

Therefore, given the described circumstances, a method performing simultaneously
  segmentation in native \gls*{dmri} space and registration of the corresponding \gls*{t1} image
  to the \gls*{dmri} data could provide the optimal solution.
Previously, joint segmentation and registration has been successfully applied to
  longitudinal object tracking \citep{paragios_level_2003} and atlas-based
  segmentation \citep{gorthi_active_2011}.
The most extended approach is integrating active contours in image registration 
  methods.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  improved an existing method \citep{yezzi_variational_2003} based on linear registration
  to the nonlinear case implementing a free-form deformation field.
\cite{droske_mumfordshah_2009} reviewed the existing techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method in 2D using levelsets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  levelset function for the registration of several active contours, in the application
  of atlas-based segmentation.

The hypothesis under this research is that the registration and segmentation
  problems in \gls*{dmri} can be solved simultaneously, increasing the geometrical
  accuracy of the process.
We propose a tool called \emph{regseg} to exploit the prior information of shapes
  extracted on \gls*{t1} images to register the anatomical reference
  to \gls*{dmri} space, implicitly segmenting these data.
The approach uses active contours without edges \citep{chan_active_2001} that evolve driving a
  free-form deformation field of B-Spline basis functions.
Optimization is performed using a descent strategy of explicit shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}, therefore \emph{regseg}
  does not implement levelsets as most of the presented methods do.
Since the nonlinear distortion is aligned with one of the imaging axis (see 
  \autoref{sec:human_connectome}), \emph{regseg} includes an anisotropic regularizer of
  the displacements field proposed by \cite{nagel_investigation_1986}.
Finally, we evaluate \emph{regseg} using an extension of our instrumentation framework
  \citep{esteban_simulationbased_2014} that simulates known and realistic distortions
  on \gls*{dmri} data.
The evaluation includes a comparison to the \gls*{t2b} correction by integrating in the framework
  an in-house implementation of the method.

### 2015-04-07 Publications/2015-NeuroImage/05-results.tex
edits from Chus

\begin{figure*}
    \centering
    \resizebox{\textwidth}{!}{%
      \begin{tikzpicture}
        \node[inner sep=10pt, draw=black!40](fig03b) at (0,7.9)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-b}};
        \node[inner sep=10pt, draw=black!40](fig03c) at (0,0)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-c}};
        \node[circle, text=black!75] at (-9.4,10.6) {\Huge \textbf{A}};
        \node[circle, text=black!75] at (-9.4,3.5) {\Huge \textbf{B}};
      \end{tikzpicture}
    }%
  \caption{A. Visual assessment of the results on the low resolution sets:
    ``gyrus'' (top-left), ``L-shape'' (top-right), ``ball'' (bottom-left),
    and ``box'' at (bottom-right).
  In yellow color, the recovered contours after registration are represented.
  Our method showed high accuracy, as it demonstrates the almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
  Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
    clustering methods, but it is successfully segmented with our method.
  B. Quantitative evaluation of registration error in terms of average Hausdorff distance of
    surfaces at low (left) and high (right) resolutions, demonstrating that the error is
    consistently below the voxel size.
    }\label{fig:phantom}
\end{figure*}
A total of 1200 experiments (4 phantom types $\times$ 150 random warpings $\times$ 2 resolutions) were
  run using the workflow given in \autoref{fig:evphantoms}.
For each experiment, the misregistration error was measured using the Hausdorff distance
  (see \autoref{sec:experiments_evaluation}) between the theoretical $\gammaset_\text{true}$ and
  the estimation done by \emph{regseg} ($\hat{\gammaset}_{test}$).
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the violin plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals,
  reported in \autoref{tab:ci_phantom}.

>>>>

\begin{figure*}
    \centering
    \resizebox{\textwidth}{!}{%
      \begin{tikzpicture}
        \node[inner sep=10pt, draw=black!40](fig03b) at (0,7.9)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-b}};
        \node[inner sep=10pt, draw=black!40](fig03c) at (0,0)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-c}};
        \node[circle, text=black!75] at (-9.4,10.6) {\Huge \textbf{A}};
        \node[circle, text=black!75] at (-9.4,3.5) {\Huge \textbf{B}};
      \end{tikzpicture}
    }%
  \caption{A. Visual assessment of the results on the low resolution sets:
    ``gyrus'' (top-left), ``L-shape'' (top-right), ``ball'' (bottom-left),
    and ``box'' at (bottom-right).
  In yellow color, the recovered contours after registration are represented.
  Our method showed high accuracy, as it demonstrates the almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
  Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
    clustering methods, but it is successfully segmented with our method.
  B. Quantitative evaluation of registration error in terms of average Hausdorff distance of
    surfaces at low (left) and high (right) resolutions, demonstrating that the error is
    consistently below the voxel size.
    }\label{fig:phantom}
\end{figure*}
A total of 1200 experiments (4 phantom types $\times$ 150 random warpings $\times$ 2 resolutions) were
  run using the workflow given in \autoref{fig:evphantoms}.
For each experiment, the misregistration error was measured using the Hausdorff distance
  (see \autoref{sec:experiments_evaluation}) between the theoretical $\gammaset_\text{true}$ and
  the estimation done by \emph{regseg} ($\hat{\gammaset}_{test}$).
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the violin plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
In order to relate average misregistration error to the resolution of the moving image,
  we proceeded as follows.
First, we confirmed that the vertex-wise error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$, for each phantom type).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals,
  reported in \autoref{tab:ci_phantom}.
